A general Nullstellensatz for generalized spaces

### 2019-03-17 paper-qcoh.tex
The coherent case

\begin{lemma}Let~$\E$ be a Grothendieck topos. Let~$\TT$ be a coherent theory
internal to~$\E$. Let~$\sigma$ be a coherent sequent over the signature
of~$\TT$. Then the following statements are equivalent.
\begin{enumerate}
\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of externally
finite shape}.$ \smallskip
\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
external shape}$, that is~$\E \models \speak{$\sigma$ is provable$^\star$
modulo~$\TT$}$. \smallskip
\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
internal shape}$, that is~$\E \models \speak{$\sigma$ is provable
modulo~$\TT$}$. \smallskip
\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of internally
finite shape}.$
\end{enumerate}
\end{lemma}

@@

% XXX coherent case

>>>>

\begin{lemma}\label{lemma:provability-in-coherent-case}
Let~$\E$ be a Grothendieck topos. Let~$\TT$ be a coherent theory
internal to~$\E$. Let~$\sigma$ be a coherent sequent over the signature
of~$\TT$. Then the following statements are equivalent.
\begin{enumerate}
\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of externally
finite shape}.$ \smallskip
\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
external shape}$, that is~$\E \models \speak{$\sigma$ is provable$^\star$
modulo~$\TT$}$. \smallskip
\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of arbitrary
internal shape}$, that is~$\E \models \speak{$\sigma$ is provable
modulo~$\TT$}$. \smallskip
\item $\E \models \speak{There is a~$\TT$-derivation of~$\sigma$ of internally
finite shape}.$
\end{enumerate}
\end{lemma}

@@

The Nullstellensatz of Theorem~\ref{thm:nullstellensatz} and the
characterization of Theorem~\ref{thm:characterization} pertain to geometric
logic and hence require some care in dealing with the flexibility of
disjunctions. In particular, the internal statement of
Theorem~\ref{thm:nullstellensatz} requires external ingredients in referring to
geometric$^\star$ sequents and provability$^\star$. A natural question is
therefore whether, in the case that~$\TT$ is a coherent theory, the statements
of these two theorems can be simplified.

As stated in Lemma~\ref{lemma:provability-in-coherent-case}, for coherent
sequents modulo a coherent theory there is no difference between provability in
geometric logic, provability$^\star$ in geometric logic and provability in
coherent logic. However, while geometric logic is powerful enough to express
provability$^\star$ in geometric logic, coherent logic is not powerful enough
to express provability in coherent logic. ``Coherent logic cannot eat itself.''
Hence the the analogue of Theorem~\ref{thm:characterization} cannot be
formulated for coherent logic.

A fragment of logic which is still finitary, but is powerful enough to express
its own provability predicate, is coherent existential fixed-point
logic~\cite{blass:topoi-and-computation,blass:existential-fixed-point-logic}.
For this fragment, we have the following characterization.

\begin{scholium}Let~$\TT$ be a coherent theory (or more generally theory in coherent
existential fixed-point logic). Let~$\alpha$ be a finitary first-order
formula over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$. \smallskip
\item The formula~$\alpha$ is provable in coherent existential fixed-point logic
modulo the axioms of~$\TT$ and the additional axioms
\begin{equation}
  \speak{$\sigma$ holds for~$U_\TT$} \Longrightarrow \speak{$\ul{\TT}/U_\TT$
  proves~$\sigma$}
\end{equation}
where~$\sigma$ ranges over the formulas of coherent existential fixed-point
logic over the signature of~$\TT$.
\end{enumerate}
\end{scholium}

\begin{proof}The proofs of Theorem~\ref{thm:nullstellensatz} and
Theorem~\ref{thm:characterization} can be adapted to the setting of coherent
existential fixed-point logic.
\end{proof}
